---
title: "4_random_fortest"
author: "Pedro"
date: "8/9/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

# 0 - Load libraries & data

```{r}

#load necessary libraries
library(phyloseq)
library(ggplot2)
library(tibble)
library(dplyr)
library(Boruta)
library(caret)
library(randomForest)
library(metacoder)
library(scorer) # several functions to evaluate regression models


#load the phyloseq objects containing all microbiome data
load("./Code/phyloseq_objects.RData")
```


# 1 - prepare phyloseq object to be an input in Boruta
```{r boruta}




#transpose phtseq otu table  
otu_cells_wide <- as.data.frame(t(otu_table(physeq_goodSamples_rarefied)))%>% 
                              rownames_to_column(var = "sample")

# extract sample data
metadata_rf <-as(sample_data(physeq_goodSamples_rarefied),"data.frame")%>%
              rownames_to_column(var = "sample")

#add the variable classification you want to predict with the random forest
Boruta_input<-
  merge(select(metadata_rf,sample, "Treatment"),
        otu_cells_wide,
        by = "sample",
        all.y = TRUE)%>%
  column_to_rownames(var = "sample")
 

#check your df
str(Boruta_input[1:10,1:10]) # if your "Treatment" is not a factor, boruta won't work
Boruta_input$Treatment<-as.factor(Boruta_input$Treatment)



```


# 2 - run Boruta to define imporntat features
```{r}
#run borura
set.seed(456987)
Boruta_output<- Boruta(Treatment~.,   
                       data = Boruta_input,
                       doTrace=2, 
                       maxRuns = 1000, 
                       ntree = 8000) #increase the maximum number of runs to decrease the number of tenttively important OTUs. increase the number of trees to increase precision. decrease either to reduce computational time.

#save the otuput, as it takes some 15 minutes to calculate
save(Boruta_output, file = "./Results/Boruta_output.RData")

```



# 3 - check boruta results/output

```{r boruta}

#let Boruta decide if tentative features are ultimatetly important or not ; 
fixed_boruta_objt<- TentativeRoughFix(Boruta_output)

# get a list of ASVs defined as inportant by Boruta
boruta_ASV_list<- getSelectedAttributes(fixed_boruta_objt)

# get the list of ASVs defined as inportant by Boruta in formula format ; this can be used to calculate precision
boruta_formula<- getConfirmedFormula(fixed_boruta_objt)

#make a plot showing imporance of features
boruta_plot<- plot(fixed_boruta_objt)

#make a plot showing imporance and classification of features over iteration time
boruta_history<-  plotImpHistory(fixed_boruta_objt)

# define ASVs tagged as important by boruta
important_ASVs<-filter(attStats(fixed_boruta_objt), decision=="Confirmed")%>%
    rownames()

#get boruta stats of ASVs confirmed to be important
rf_importance_byOTU<-filter(attStats(fixed_boruta_objt), decision=="Confirmed")%>%
    rownames_to_column(var = "otu_id")

```

# 4 - calculate precision and kappa with confusion matrixes
```{r}



# define random seed
set.seed(4551)

#define training model
train.control <- trainControl(method = "repeatedcv", # set trainig/data split controls for the train function
                              number = 5, 
                              repeats = 200)
# calculate precision
model_borutized <- train(boruta_formula, 
                         data = Boruta_input, 
                         method = "rf", #execute training based on random forest; model is based on borut formula of important features
               trControl = train.control, ntree=1000)

# plot confusion matrix
confusionMatrix(model_borutized)

# save teh output, as it takes ~30 min to calculate
save(model_borutized, file = "./Results/model_borutized.RData")

```

## 5 -Visualize the features on a heat tree matrix

Just for the record, these are teh ASVs confirmed as importat by boruta:

Treatment ~ ASV_2 + ASV_3 + ASV_5 + ASV_6 + ASV_9 + ASV_10 + 
    ASV_12 + ASV_13 + ASV_15 + ASV_18 + ASV_19 + ASV_23 + ASV_26 + 
    ASV_28 + ASV_30 + ASV_36 + ASV_37 + ASV_38 + ASV_40 + ASV_44 + 
    ASV_47 + ASV_51 + ASV_65 + ASV_72 + ASV_81 + ASV_82 + ASV_90 + 
    ASV_94 + ASV_97 + ASV_108 + ASV_120 + ASV_122 + ASV_127 + 
    ASV_134 + ASV_142 + ASV_149 + ASV_154 + ASV_158 + ASV_161 + 
    ASV_166 + ASV_172 + ASV_173 + ASV_176 + ASV_204 + ASV_223 + 
    ASV_254 + ASV_261 + ASV_266 + ASV_268 + ASV_283 + ASV_294 + 
    ASV_309 + ASV_341 + ASV_342 + ASV_346 + ASV_353 + ASV_365 + 
    ASV_377 + ASV_384 + ASV_391 + ASV_399 + ASV_406 + ASV_433 + 
    ASV_434 + ASV_442 + ASV_449 + ASV_457 + ASV_458 + ASV_511 + 
    ASV_516 + ASV_518 + ASV_521 + ASV_549 + ASV_588 + ASV_629 + 
    ASV_646 + ASV_660 + ASV_688 + ASV_710 + ASV_729 + ASV_773 + 
    ASV_797 + ASV_808 + ASV_832 + ASV_858 + ASV_961 + ASV_1026 + 
    ASV_1094 + ASV_1097 + ASV_1110 + ASV_1215 + ASV_1516 + ASV_1570 + 
    ASV_1708

```{r}



# first, get only the ASVs detected as differentially abundant by Deseq2
rf_important_heatTrees<-prune_taxa(taxa = important_ASVs, physeq_goodSamples_rarefied)


#adds ASV IDs as the "species"
tax_table(rf_important_heatTrees)[,7]<-important_ASVs

#remove unecessary taxonomic indo (dada2id, "S__" and" above_selected)
tax_table(rf_important_heatTrees )<-tax_table(rf_important_heatTrees)[,1:7]



# let's remove the "r__"ranks from the taxonomy, they can be useful but will polute our plot
tax_table(rf_important_heatTrees )[, colnames(tax_table(rf_important_heatTrees ))] <- gsub(tax_table(rf_important_heatTrees )[, colnames(tax_table(rf_important_heatTrees ))],     pattern = "[a-z]__", replacement = "")

# transform from phyloseq to  taxmap object
rf_important_metacoder<-parse_phyloseq(rf_important_heatTrees )

#get abundance per taxon
rf_important_metacoder$data$tax_abund<-calc_taxon_abund(obj = rf_important_metacoder, 
                                      data = "otu_table",
                                      cols = rf_important_metacoder$data$sample_data$sample_id)
#get occuence of per sample type
rf_important_metacoder$data$tax_occ <- calc_n_samples(obj = rf_important_metacoder, 
                                    data = "tax_abund", 
                                    groups = rf_important_metacoder$data$sample_data$Treatment, 
                                    cols = rf_important_metacoder$data$sample_data$sample_id)


############################## now, let's plot a matrix heat tree for the MeJA comparisons


rf_important_metacoder$data$diff_table <- compare_groups(obj = rf_important_metacoder,
                                      dataset = "tax_abund",
                                      cols = rf_important_metacoder$data$sample_data$sample_id, # What columns of sample data to use
                                      groups = rf_important_metacoder$data$sample_data$Treatment) # What category each sample is assigned to

# set differental log ratio to 0 based on adjusted p values
rf_important_metacoder$data$diff_table$log2_median_ratio[rf_important_metacoder$data$diff_table$wilcox_p_value > 0.05] <- 0

length(rf_important_metacoder$data$diff_table$log2_median_ratio)
#check this object to find terminal leaves that are consistent across treatments, then...
# highlight a taxonomic rank with lots of members being differentially selected
rf_important_metacoder$data$diff_table$treatment_1
tail(rf_important_metacoder$data$otu_table)









########### let's add RF importance to the metacoder object######

# add metacoder taxon ID to RF importnat OTUs
rf_importance_byOTU$taxon_id<-rf_important_metacoder$data$otu_table$taxon_id

# add RF importance to metacoder object
rf_important_metacoder$data$RF_importance<-tibble(rf_importance_byOTU[,c(1:2,8)])

#get summed importance per taxon
rf_important_metacoder$data$RF_importance_nodes<-calc_taxon_abund(obj = rf_important_metacoder, 
                                      data = "RF_importance")

#define number of subtaxa
subtaxa_number<-n_subtaxa(rf_important_metacoder)+1

# transform zeros into 1 so you don't try to divide by zero on the next step
subtaxa_number[subtaxa_number==0]<-1

# add this subtaxa number to part of the metacoder tree object
rf_important_metacoder$data$RF_importance_nodes$subtaxa_number<-subtaxa_number

# divide the summed meam importance y the number of subtaxa. ASV nodes shuold ahve the same importance as the original object
rf_important_metacoder$data$RF_importance_nodes$mean_meamImp_per_taxa<-
  rf_important_metacoder$data$RF_importance_nodes$meanImp /
  rf_important_metacoder$data$RF_importance_nodes$subtaxa_number

#define the minimum and maximum ASV importance to plot
min_RF_importance<-min(rf_importance_byOTU$meanImp) 
max_RF_importance<-max(rf_importance_byOTU$meanImp) 


########### done! ######












#plot matrix tree
set.seed(1)
heat_tree_matrix(rf_important_metacoder,
                 data = "diff_table",
                 node_size = rf_important_metacoder$data$RF_importance_nodes$mean_meamImp_per_taxa, # n_obs is a function that calculates, in this case, the number of OTUs per taxon
                 node_label = taxon_names,
                 node_color = log2_median_ratio, # A column from `obj$data$diff_table`
                 node_color_range = diverging_palette(), # The built-in palette for diverging data
                 node_color_trans = "linear", # The default is scaled by circle area
                 node_color_interval = c(-3, 3), # The range of `log2_median_ratio` to display
                 edge_color_interval = c(-3, 3), # The range of `log2_median_ratio` to display
                 overlap_avoidance = 1.5,
                node_size_interval = c(min_RF_importance, max_RF_importance), # define node size range according RF importances
                 node_label_size_range = c(0.005, 0.015),
                 node_size_range = c(0.009, 0.03),
                 label_small_trees = TRUE,
                 node_size_axis_label = "Size: Number of OTUs",
                 node_color_axis_label = "Color: Log2 ratio median proportions",
                 layout = "davidson-harel", # The primary layout algorithm
                 initial_layout = "reingold-tilford", # The layout algorithm that initializes node locations
                 output_file = "./Results/RF_features_heat_tree.pdf") # Saves the plot as a pdf file


```





##5b - barplot of rf features
```{r}




#get boruta stats of ASVs confirmed to be important
rf_importance_byOTU<-filter(attStats(fixed_boruta_objt), decision=="Confirmed")%>%
    rownames_to_column(var = "OTU")

#sort ASVs by importance
rf_importance_byOTU<-rf_importance_byOTU[order(rf_importance_byOTU$meanImp, decreasing = TRUE),]


#turn ps object into a dataframe
melted_ps<- psmelt(rf_important_heatTrees)

 # join melted ps object created above with the boruta stats
 rf_ASVs_df<-left_join(melted_ps,rf_importance_byOTU)
 

rf_ASVs_df$OTU<-base::factor(rf_ASVs_df$OTU, levels = c(rf_importance_byOTU$OTU))

# quick plot bar with relative abudances, for a quick overview of dity data

ggplot(data=rf_ASVs_df, aes(x=OTU, y =log(Abundance)))+
  geom_boxplot(aes())+
  scale_fill_hue()+
  theme_bw()+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 


ggplot(data=rf_ASVs_df, aes(x=Genus, y =medianImp ))+
  geom_boxplot(aes())+
  scale_fill_hue()+
  theme_bw()+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 

table(rf_ASVs_df)



```




# 6 - run boruta to predict flowers
```{r}


#add the variable classification you want to predict with the random forest
Boruta_input_flower<-
  merge(select(metadata_rf,sample, "Numberofflowers"),
        otu_cells_wide,
        by = "sample",
        all.y = TRUE)%>%
  column_to_rownames(var = "sample")
 

#check your df
str(Boruta_input_flower[1:10,1:10]) # if your "Numberofflowers" is not numeric, boruta won't work
Boruta_input_flower$Numberofflowers<-as.numeric(Boruta_input_flower$Numberofflowers)



#run borura
set.seed(456987)
Boruta_output_flower<- Boruta(Numberofflowers~.,   
                       data = Boruta_input_flower,
                       doTrace=2, 
                       maxRuns = 1000, 
                       ntree = 8000) #increase the maximum number of runs to decrease the number of tenttively important OTUs. increase the number of trees to increase precision. decrease either to reduce computational time.


#let Boruta decide if tentative features are ultimatetly important or not ; 
fixed_boruta_flower_objt<- TentativeRoughFix(Boruta_output_flower)

# get a list of ASVs defined as inportant by Boruta
boruta_flower_ASV_list<- getSelectedAttributes(fixed_boruta_flower_objt)

# get the list of ASVs defined as inportant by Boruta in formula format ; this can be used to calculate precision
boruta_flower_formula<- getConfirmedFormula(fixed_boruta_flower_objt)

# save teh output, as it takes ~10 min to calculate
save(Boruta_output_flower, file = "./Results/Boruta_output_flower.RData")

# define ASVs tagged as important by boruta
important_flower_ASVs<-filter(attStats(fixed_boruta_flower_objt), decision=="Confirmed")%>%
    rownames()

#get boruta stats of ASVs confirmed to be important
rf_importance_flower_byOTU<-filter(attStats(fixed_boruta_flower_objt), decision=="Confirmed")%>%
    rownames_to_column(var = "otu_id")




```




# 7 - Calcualte precisiona and kappa for boruta flowers


```{r}



# define random seed
set.seed(4551)

#define training model
train.control <- trainControl(method = "repeatedcv", # set trainig/data split controls for the train function
                              number = 5, 
                              repeats = 200)
# calculate precision
model_borutized_flower <- train(boruta_flower_formula, 
                         data = Boruta_input_flower, 
                         method = "rf", #execute training based on random forest; model is based on borut formula of important features
               trControl = train.control, ntree=1000)

# check mean_squared_error
mean_squared_error(y_true = Boruta_input_flower$Numberofflowers,
                     y_pred = predict(model_borutized_flower, Boruta_input_flower[,-1]))

hist(squared_error(y_true = Boruta_input_flower$Numberofflowers,
                     y_pred = predict(model_borutized_flower, Boruta_input_flower[,-1])))

# check root mean_squared_error
sqrt(mean_squared_error(y_true = Boruta_input_flower$Numberofflowers,
                     y_pred = predict(model_borutized_flower, Boruta_input_flower[,-1])))

#check explained variance
r2_score(y_true = Boruta_input_flower$Numberofflowers,
                     y_pred = predict(model_borutized_flower, Boruta_input_flower[,-1]))


# visualize predicitons VS real data
plot(Boruta_input_flower$Numberofflowers, predict(model_borutized_flower, Boruta_input_flower[,-1]))








# save teh output, as it takes ~30 min to calculate
save(model_borutized_flower, file = "./Results/model_borutized_flower.RData")

```


## 8 -  Visualize the flower features on a heat tree 

```{r}
# first, get only the ASVs detected as differentially abundant by Deseq2
rf_important_heatTrees<-prune_taxa(taxa = important_ASVs, physeq_goodSamples_rarefied)

#adds ASV IDs as the "species"
tax_table(rf_important_heatTrees)[,7]<-important_ASVs

#remove unecessary taxonomic indo (dada2id, "S__" and" above_selected)
tax_table(rf_important_heatTrees )<-tax_table(rf_important_heatTrees)[,1:7]



# let's remove the "r__"ranks from the taxonomy, they can be useful but will polute our plot
tax_table(rf_important_heatTrees )[, colnames(tax_table(rf_important_heatTrees ))] <- gsub(tax_table(rf_important_heatTrees )[, colnames(tax_table(rf_important_heatTrees ))],     pattern = "[a-z]__", replacement = "")


####################### now make it a metacoder object

# transform from phyloseq to  taxmap object
rf_important_metacoder<-parse_phyloseq(rf_important_heatTrees )



#get abundance per taxon
rf_important_metacoder$data$tax_abund<-calc_taxon_abund(obj = rf_important_metacoder, 
                                      data = "otu_table",
                                      cols = rf_important_metacoder$data$sample_data$sample_id)

#get mean taxon abudance per group
rf_important_metacoder$data$compare_group<-calc_group_mean(rf_important_metacoder, 
                                                      "tax_abund",
                                              rf_important_metacoder$data$sample_data$Treatment)

#get occuence of per sample type
rf_important_metacoder$data$tax_occ <- calc_n_samples(obj = rf_important_metacoder, 
                                    data = "tax_abund", 
                                    groups = rf_important_metacoder$data$sample_data$Treatment, 
                                    cols = rf_important_metacoder$data$sample_data$sample_id)






########### let's add RF importance to the metacoder object######

# add metacoder taxon ID to RF importnat OTUs
rf_importance_byOTU$taxon_id<-rf_important_metacoder$data$otu_table$taxon_id

# add RF importance to metacoder object
rf_important_metacoder$data$RF_importance<-tibble(rf_importance_byOTU[,c(1:2,8)])

#get summed importance per taxon
rf_important_metacoder$data$RF_importance_nodes<-calc_taxon_abund(obj = rf_important_metacoder, 
                                      data = "RF_importance")

#define number of subtaxa
subtaxa_number<-n_subtaxa(rf_important_metacoder)+1

# transform zeros into 1 so you don't try to divide by zero on the next step
subtaxa_number[subtaxa_number==0]<-1

# add this subtaxa number to part of the metacoder tree object
rf_important_metacoder$data$RF_importance_nodes$subtaxa_number<-subtaxa_number

# divide the summed meam importance y the number of subtaxa. ASV nodes shuold ahve the same importance as the original object
rf_important_metacoder$data$RF_importance_nodes$mean_meamImp_per_taxa<-
  rf_important_metacoder$data$RF_importance_nodes$meanImp /
  rf_important_metacoder$data$RF_importance_nodes$subtaxa_number

#define the minimum and maximum ASV importance to plot
min_RF_importance<-min(rf_importance_byOTU$meanImp) 
max_RF_importance<-max(rf_importance_byOTU$meanImp) 

########### done! ######








#plot heat tree
set.seed(1)
heat_tree(rf_important_metacoder,
                 node_size = rf_important_metacoder$data$RF_importance_nodes$mean_meamImp_per_taxa, # n_obs is a function that calculates, in this case, the number of OTUs per taxon
                 node_label = taxon_names,
                 node_color = rowSums(rf_important_metacoder$data$tax_abund[,c(-1)]), # A column from `obj$data$diff_table`
                node_size_interval = c(min_RF_importance, max_RF_importance), # define node size range according RF importances
                 node_label_size_range = c(0.005, 0.015),
                 node_size_range = c(0.009, 0.03),
                 node_size_axis_label = "Size: RF importance OTU count",
                 node_color_axis_label = "Color: Number of sequences",
                 layout = "davidson-harel", # The primary layout algorithm
                 initial_layout = "reingold-tilford", # The layout algorithm that initializes node locations
                 output_file = "./Results/RF_features_flowers.pdf") # Saves the plot as a pdf file




```






# 9 - Run boruta to predict plant weight

There are no imporntat atributes for plant weight! ASVs cannot determine planr biomass with Boruta RF 
```{r}


#add the variable classification you want to predict with the random forest
Boruta_input_biomass<-
  merge(select(metadata_rf,sample, "Fressabovegroundbiomass_g"),
        otu_cells_wide,
        by = "sample",
        all.y = TRUE)%>%
  column_to_rownames(var = "sample")
 

#check your df
str(Boruta_input_biomass[1:10,1:10]) # if your "Fressabovegroundbiomass_g" is not numeric, boruta won't work
Boruta_input_biomass$Fressabovegroundbiomass_g<-as.numeric(Boruta_input_biomass$Fressabovegroundbiomass_g)



#run borura
set.seed(456987)
Boruta_output_biomass<- Boruta(Fressabovegroundbiomass_g~.,   
                       data = Boruta_input_biomass,
                       doTrace=2, 
                       maxRuns = 1000, 
                       ntree = 20000) #increase the maximum number of runs to decrease the number of tenttively important OTUs. increase the number of trees to increase precision. decrease either to reduce computational time.


```



#10 - check overlaps  in the highlighted data
```{r}

# are differentially abudnant ASVs also RF importnat ASVs? 
boruta_ASV_list[boruta_ASV_list %in% differentialy_abundant_ASVs]

# which ASVs are RF-important both for treatments and number of flowers, making it a "bi-important" ASV? 
bi_important_ASVs<-important_flower_ASVs[important_flower_ASVs %in% boruta_ASV_list]

#what are the taxonomies of these ASVs? detected as "bi-importat"
tax_table(prune_taxa(bi_important_ASVs, physeq_goodSamples_rarefied))

#what is the RF mdoel importance for these "bi-important" ASVs?
<<<<<<< HEAD
filter(rf_importance_flower_byOTU, otu_id %in% bi_important_ASVs)
filter(rf_importance_byOTU, otu_id %in% bi_important_ASVs)
=======
>>>>>>> 3f2bae0dc68f9b55461b0660de37c73af8d2df75


```

